\doctitle

### 2023-05-02 pinode.tex
Add sensitivity+scalability analyses and covid example
Incorporate edits after submitting the rebuttal

\section{Introduction}
\edit{Write a big-picture intro to machine learning for physics}
\edit{Replace mentions of paper with work}
\subsection{Related Work}
\edit{Split original intro into prior works paragraphs: DL for Ph, ROMs, PIML}
\edit{Add physics-informed ML overview (PINN, DeepONet, NFO)}
        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}. At the moment of writing this paper such systems are incapable of real-time complex simulations, but they already can run low-dimensional pre-trained models, thus inviting the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation} . ROMs, therefore, are essential for enabling design optimization, uncertainty propagation, predictive models, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

        To be suitable for control, a ROM training method needs to find a low-dimensional manifold and a dynamics on it that together yield both high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}. Most traditional ROMs are projection-based, e.g. dynamic mode decomposition (DMD) \cite{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \cite{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace. This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of basis functions (spatial modes) \cite{benner2015survey}. 
        One challenge for POD methods is their intrusive nature, i.e. requiring access to the solver codes.
       To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and then exploit lifting to fit the latent space dynamics data into
        polynomial, typically quadratic, models.
        These models, however, are (i) limited in representation power (up to quadratic, e.g. for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.

        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \cite{lee2020model,gin2021deep,champion2019data,kim2019deep}. In practice, however, they require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system. Large demand for data significantly limits the use of such models in physics applications where the data can be expensive to obtain.

        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \cite{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs. For example, in HVAC applications, one may collect data from a room with two windows but not from one room for every possible number of windows. In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \cite{nabi2020improving}. In such situations embedding the knowledge of physics into a model becomes necessary to improve extrapolation performance, and for which several approaches have recently been proposed.
        For instance, the seminal works \cite{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression. Recently, Cranmer et al. \cite{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations. They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution- data better than the GNN itself. However, symbolic regression also suffers from excessive computational costs,  and may be prone to overfitting.

        Another example of incorporating physics in ROMs is the use of parametric models at the latent space, e.g. by using the sparse identification of nonlinear dynamics (SINDy) \cite{brunton2016discovering,champion2019data}. For instance, \cite{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics. However, such loss is highly sensitive to noise in data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}. 
        Collocation-based enforcement of the physics, i.e. projection of the candid functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties. Recently, Liu et al. \cite{liu2022physics} used auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM. However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \cite{page2019koopman}.
        Finally, recent works on NeuralODE (NODE) \cite{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g. a network) as a latent space dynamics model, significantly extending the set of models for the latent dynamics that one can train efficiently.

        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE at the latent space to model complex and nonlinear dynamics. Our goal is to reduce their demand for data and improve overall forecasting stability under challenging training conditions. To achieve that we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2. Then, in Section 3, we show that addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.
        \edit{Add Koopman example to the intro}

@@

            \begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/graphical-abstract.png}
    \caption{We propose utilizing a collocation points technique from numerical analysis to transfer knowledge of physics into continuous-time reduced-order models (ROMs). Such physics-informed models yield orders of magnitude more accurate predictions in tasks of far-out forecasting, predictions for out-of-distribution initial conditions, and learning from high-noise data. Good performance on those tasks is crucial for using ROMs in problems of compressive sensing and control.}
    \label{fig:architecture}
\end{figure}

\section{Method}
\label{sec:method}
\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics}
We consider an autonomous dynamical system on a finite space $\XX \subseteq \mathbb{R}^n$

\begin{equation}
    \label{eq:generic_stationary_ode}
        \ddt\bd{x}(t) = \bd{f}(\bd{x}(t))
\end{equation}

In real-world applications it is often expensive to solve the relationship~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional. However, a variety of works provided both theoretical \cite{holmes2012turbulence} and practical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$. In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:

@@

We call the space $\XX$ an observable space, and $\ZZ$ a latent space. When an invertible mapping $\psi:\ \ZZ \to \XX$ between the observable and the latent spaces is known, one can predict the dynamics of the system $\bd{x}$ at a future time $T$ by projecting the initial condition $\bd{x}(0)$ into the latent space, performing an integration there, and mapping the resulting trajectory back to the observable space:
\begin{equation}
\label{eq:integration_in_latent_space}
\begin{split}
    \bd{z}(0) & = \psi^{-1}(\bd{x}(0)) \\
    \bd{z}(T) & = \bd{z}(0) + \int_{0}^T\bd{h}(\bd{z}(t))dt \\
    \bd{x}(T) & = \psi(\bd{z}(T))
\end{split}
\end{equation}

When $m << n$ we refer to the triplet $(\psi, \psi^{-1}, \bd{h})$ as a Reduced-Order Model (ROM) of $\bd{f}$. It is often the case that for a given system $\bd{f}$ there exists no ROM $(\psi, \psi^{-1}, \bd{h})$ such that the relation~(\ref{eq:integration_in_latent_space}) holds exactly. In this case, we seek an \textit{approximation} ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ that minimizes the difference between the data $x(t)$ and the prediction $\hat{x}(t)$ over a chosen class of models $(\psi_\theta, \phi_\theta, h_\theta)$ parameterized by $\theta$.

Multiple real-world applications necessitate using ROMs instead of integrating the relation~(\ref{eq:generic_stationary_ode}) directly. For example, integrating~(\ref{eq:generic_stationary_ode}) may be computationally intractable especially on platforms with limited computing capability such as embedded and autonomous devices. For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stocks equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances. On the other hand, integrating~(\ref{eq:integration_in_latent_space}) may be cheap when $m << n$. Finally, even when solving~(\ref{eq:generic_stationary_ode}) is possible in real time (e.g. by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable. Indeed, executing control requires \textit{multiple} iterative evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.

\paragraph{Architecture} In this work we model $\psi$, $\psi^{-1}$, and $\bd{h}$ with fully-connected neural networks $\psi_\theta$, $\phi_\theta$, and $h_\theta$, respectively. Specifically, the pair ($\psi$, $\psi^{-1}$) is modeled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modeled with a fully-connected network $h_\theta$. Figure~\ref{fig:architecture} visualizes the architecture of the model.

\paragraph{Data-Driven Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses. The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data. Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation without loss of generality, in the rest of the paper we assume that all trajectories were recorded over the same time-frame with an equal and uniform step-size.

\paragraph{Physics-Informed Loss} In their recent work, \cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics. To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
\begin{equation}
    \label{eq:chain_rule_1}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bf{f}(\bd{x(t)})
\end{equation}
On the other hand, by the definition of $\psi$ and $\bd{h}$ we have that
\begin{equation}
    \label{eq:chain_rule_2}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \bd{h}(\phi(\bd{x}(t))
\end{equation}
Combining Equations~(\ref{eq:chain_rule_1}) and~(\ref{eq:chain_rule_2}) we get that
\begin{equation}
    \label{eq:physsics_informed_equation}
    \bd{h}(\phi(\bd{x}(t)) = \nabla\phi(\bd{x})^T\bd{f}(\bd{x})
\end{equation}
Equation~(\ref{eq:physsics_informed_equation}) links the dynamics $\bd{h}(\bd{z})$ and the encoder $\phi(\bd{x})$ with the known equation $\bd{f}(\bd{x})$ and is true for all $z \in \ZZ$ and $x \in \XX$. Hence, knowledge of $\bd{f}$ can be assimilated into the model by evaluating Equation~(\ref{eq:physsics_informed_equation}) on a set of $N$ carefully sampled points $\bar{\bd{x}}_i \in \XX$, $i \in [1, \dots, N]$:

@@

We refer to the points $\bar{\bd{x}}_i$ as \textit{collocations}.

\paragraph{Collocations}
\label{sec:collocations_conditions} We define a collocation as pair $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$. Collocations are samples from the space $\XX \times Im_{f}(\XX)$, and they should satisfy three conditions, ordered by importance:
\begin{enumerate}
    \item \textbf{Simplicity}: $\bd{f}(\bar{\bd{x}}_j)$ should be computationally cheap to evaluate. It is especially important for PDE systems, where $\bd{f}$ may involve high-order derivatives.
    \item \textbf{Representativeness}: $\bar{\bd{x}}_j$ should cover the space of states where one aims to improve the model's performance or stability. Collocation points that a model might encounter and that are not represented by data snapshots are the best candidates.
    \item \textbf{Feasibility}: $\bar{\bd{x}}_j \in \XX$. In other words, $x_j$ should be an attainable state of the system. Collocation points outside of $\XX$ may downgrade the performance of the auto-encoder by forcing it to be an invertible function on a domain outside of $\XX$ where a true mapping $\psi$ operates on.
\end{enumerate}
Thus, an optimal sampling procedure for collocations $\bd{\bar{x}}_j$ is domain-specific and should be designed given a particular system $\bd{f}$ and available data $\bd{x}_i$. We show examples of how these conditions can be implemented for real systems in the following sections.

Our definition of collocation points follows recent works\cite{liu2022physics}, which is a further development of the original definition \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.

\paragraph{Combined Loss Function} We train the model by optimizing a sum of the physics-informed loss~(\ref{eq:loss_physics_informed}) and the data-driven loss~(\ref{eq:loss_data_driven}):
\begin{equation}
    \label{eq:loss_combined}
    \min_\theta \left[\LL^{physics}_\theta + \LL^{data}_\theta\right]  
\end{equation}
When $\omega_1 = \omega_2 = 0$ we have that $\LL^{data}_\theta = 0$, so we say that the model is (purely) \textit{\textbf{Physics-Informed}}. Similarly, when $\omega_3 = \omega_4 = 0$ we have that $\LL^{physics}_\theta = 0$ and we say that the model is (purely) \textit{\textbf{Data-Driven}}. When $\omega_i \neq 0, \, \forall i$, we say that the model is \textit{\textbf{Hybrid}}.

We use a \texttt{pytorch}~\cite{NEURIPS2019_9015} implementation of Adam algorithm~\cite{kingma2014adam} for optimization. To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\cite{chen2018neural} -- a \texttt{pytorch}-compatible implementation of Neural ODE framework.

To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), auto-encoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}). Thus, we call our framework \textit{Physics-Informed Neural ODE}, or PINODE.

>>>>

\section{Introduction}
\edit{Write a big-picture intro to machine learning for physics}
\edit{Replace mentions of paper with work}
\subsection{Related Work}
\edit{Split original intro into prior works paragraphs: DL for Ph, ROMs, PIML}
\edit{Add physics-informed ML overview (PINN, DeepONet, NFO)}
        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}. At the time of writing this paper, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}. Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

        In order to enable control of high dimensional dynamical systems, a ROM training method needs to  identify a low-dimensional manifold along with dynamics on the manifold that together yield high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}. Most traditional ROMs are projection-based, e.g. dynamic mode decomposition (DMD) \cite{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \cite{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace. This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of the basis functions (spatial modes) \cite{benner2015survey}. One challenge for POD methods is their intrusive nature, i.e. requiring access to the solver codes. To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and  exploit lifting to fit the latent space dynamics data into polynomial, typically quadratic, models. These models, however, are (i) limited in representation power (up to quadratic, e.g. for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.

        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \cite{lee2020model,gin2021deep,champion2019data,kim2019deep}. In practice, however, autoencoder-based ROMs require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system. Moreover, the large demand for training data significantly limits the use of such models in physics applications where the data can be expensive to obtain.

        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \cite{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs. For example, in HVAC applications, one may collect data from a room with two windows but not from a room for every possible number of windows. In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \cite{nabi2020improving}. In such situations embedding the knowledge of physics into a model becomes necessary to improve the extrapolation performance, and for which several approaches have recently been proposed.
        For instance, the seminal works \cite{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression. Recently, Cranmer et al. \cite{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations. They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution data better than the GNN itself. However, symbolic regression also suffers from excessive computational costs,  and may be prone to overfitting.

        Another example of incorporating physics in ROMs is the use of parametric models at the latent space, e.g. by using the sparse identification of nonlinear dynamics (SINDy) \cite{brunton2016discovering,champion2019data}. For instance, \cite{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics. However, such loss is highly sensitive to noise in the data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}. 
        Collocation-based enforcement of the physics, i.e. projection of the candidate functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties. Recently, Liu et al. \cite{liu2022physics} used an auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM. However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \cite{page2019koopman}.
        Finally, recent works on NeuralODE (NODE) \cite{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g. a network) as a latent space dynamics model, significantly extending the set of models for the latent dynamics that one can train efficiently.

        In this paper, we employ autoencoders to perform nonlinear model reduction along with NODE in the latent space to model complex and nonlinear dynamics. We choose Neural ODEs in the latent space dynamics representation because of their ability to model highly non-linear dynamics, which is especially important when applications limit the size of the latent space dimension. Our goal is to reduce the demand for training data and improve the overall forecasting stability under challenging training conditions. To that end, we build on ideas from classical collocation methods of numerical analysis to embed knowledge from a known governing equation into the latent-space dynamics of a ROM, as described in Section 2. In Section 3, we show that the addition of our physics-informed loss allows for exceptional data supply strategies that improves the performance of ROMs in data-scarce settings, where training high-quality data-driven models is impossible. We demonstrate that such an approach not only reduces the need for large training data-sets and produces highly-accurate and long-term stable models, but also leads to the discovery of more compact latent spaces, which is especially important for applications in compressed sensing and control.

        \edit{Add Koopman example to the intro}

@@

%\begin{figure}[ht]
%    \centering
%    \includegraphics[width=\textwidth]{figures/graphical-abstract.png}
%    \caption{We propose utilizing a collocation points technique from numerical analysis to transfer knowledge of physics into continuous-time reduced-order models (ROMs). Such physics-informed models yield orders of magnitude more accurate predictions in tasks of far-out forecasting, predictions for out-of-distribution initial conditions, and learning from high-noise data. Good performance on those tasks is crucial for using ROMs in problems of compressive sensing and control.}
%    \label{fig:architecture}
%\end{figure}

\section{Methods}
\label{sec:method}
\paragraph{Reduced-Order Model with Non-Linear Latent Dynamics}
We consider an autonomous dynamical system on a finite space $\XX \subseteq \mathbb{R}^n$ given by

\begin{equation}
    \label{eq:generic_stationary_ode}
        \ddt\bd{x}(t) = \bd{f}(\bd{x}(t)).
\end{equation}

In real-world applications, it is often expensive to solve equation~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional. However, a variety of works provided both theoretical \cite{holmes2012turbulence} and empirical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$. In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:

@@

We call the space $\XX$ an observable space, and $\ZZ$ a latent space. When an invertible mapping $\psi:\ \ZZ \to \XX$ between the observable and the latent spaces is known, one can predict the dynamics of the system $\bd{x}$ at a future time $T$ by projecting the initial condition $\bd{x}(0)$ into the latent space, integrating the dynamics in the latent space, and mapping the resulting trajectory back to the observable space:
\begin{equation}
\label{eq:integration_in_latent_space}
\begin{split}
    \bd{z}(0) & = \psi^{-1}(\bd{x}(0)) \\
    \bd{z}(T) & = \bd{z}(0) + \int_{0}^T\bd{h}(\bd{z}(t))dt \\
    \bd{x}(T) & = \psi(\bd{z}(T))
\end{split}
\end{equation}

When $m << n$ we refer to the triplet $(\psi, \psi^{-1}, \bd{h})$ as a Reduced-Order Model (ROM) of $\bd{f}$. It is often the case that for a given system $\bd{f}$, there exists no ROM $(\psi, \psi^{-1}, \bd{h})$ such that the relation~(\ref{eq:integration_in_latent_space}) holds exactly. In this case, we seek an \textit{approximation} ROM $(\psi_{\theta^*}, \phi_{\theta^*}, h_{\theta^*})$ that minimizes the difference between the data $x(t)$ and the prediction $\hat{x}(t)$ over a chosen class of models $(\psi_\theta, \phi_\theta, h_\theta)$ parameterized by $\theta$.

Multiple real-world applications necessitate using ROMs instead of integrating the relation~(\ref{eq:generic_stationary_ode}) directly. For example, integrating~(\ref{eq:generic_stationary_ode}) may be computationally intractable especially on platforms with limited computing capability such as embedded and autonomous devices. For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stokes equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances. On the other hand, integrating~(\ref{eq:integration_in_latent_space}) may be cheap when $m << n$. Finally, even when solving~(\ref{eq:generic_stationary_ode}) is possible in real time (e.g. by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable. Indeed, executing control requires \textit{multiple} evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/abstract_data_driven.png}
    \caption{Illustration of the autoencoer structure with neural ODE in the latent space. The data-driven part of the loss function aims to minimize a sum of two objectives: the prediction loss and the reconstruction loss. The prediction loss minimizes the difference between the data trajectories and their model predictions to ensure temporal consistency of the latent space dynamics. The reconstruction loss ensures accurate reconstruction of individual snapshots, ensuring that the autoencoder behaves as an invertible mapping on all snapshots.}
    \label{fig:data_driven_loss}
\end{figure}

\paragraph{Architecture} In this work we model $\psi$, $\psi^{-1}$, and $\bd{h}$ with fully-connected neural networks $\psi_\theta$, $\phi_\theta$, and $h_\theta$, respectively. Specifically, the pair ($\psi$, $\psi^{-1}$) is modelled with an auto-encoder $(\psi_\theta, \phi_\theta)$, and $\bd{h}$ is modelled with a fully-connected network $h_\theta$. Figure~\ref{fig:data_driven_loss} visualizes the architecture of the model.

\paragraph{Data-Driven Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses. The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data, as illustrated on Figure~\ref{fig:data_driven_loss}.

 Formally, for a given set of trajectories $\bd{x}_i$, $i \in [1 \dots k]$, where each trajectory $\bd{x}_i \in \mathbb{R}^{n \times p}$ is a set of $p$ snapshots that correspond to the recorded states of the system for $p$ time-steps, $t_j$, $j \in [1, \dots, p]$, the loss function $\LL^{data}_\theta$ is defined as:

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation, without loss of generality, in the rest of the paper we assume that all trajectories are recorded over the same time-frame with the same uniform step-size. To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}). Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited. This is especially useful in applications where the size of the latent space dimension needs to be small~\cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.

\paragraph{Physics-Informed Loss} In their recent work, Liu et al.~\cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics. To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
\begin{equation}
    \label{eq:chain_rule_1}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bd{f}(\bd{x(t)})
\end{equation}
On the other hand, by the definition of $\psi$ and $\bd{h}$ we have that
\begin{equation}
    \label{eq:chain_rule_2}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \bd{h}(\phi(\bd{x}(t))
\end{equation}
Combining Equations~(\ref{eq:chain_rule_1}) and~(\ref{eq:chain_rule_2}) we get that
\begin{equation}
    \label{eq:physsics_informed_equation}
    \bd{h}(\phi(\bd{x}(t)) = \nabla\phi(\bd{x})^T\bd{f}(\bd{x})
\end{equation}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/abstract_physics_informed.png}
    \caption{The physics-informed loss function compares gradient fields in the current latent space with what a correctly-learned field should be in this latent space on set of collocation points.}
    \label{fig:physics_informed_loss}
\end{figure}

Equation~(\ref{eq:physsics_informed_equation}) links the dynamics $\bd{h}(\bd{z})$ and the encoder $\phi(\bd{x})$ with the known equation $\bd{f}(\bd{x})$ and is true for all $z \in \ZZ$ and $x \in \XX$. Hence, as shown on Figure~\ref{fig:physics_informed_loss}, knowledge of $\bd{f}$ can be assimilated into the model by evaluating Equation~(\ref{eq:physsics_informed_equation}) on a set of $N$ carefully sampled points $\bar{\bd{x}}_i \in \XX$, $i \in [1, \dots, N]$:

@@

We refer to the points $\bar{\bd{x}}_i$ as \textit{collocation points}.

\paragraph{Collocation Points}
\label{sec:collocations_conditions} We define a collocation as pair $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$. collocation points are samples from the space $\XX \times Im_{f}(\XX)$, and they should satisfy three conditions, ordered by importance:
\begin{enumerate}
    \item \textbf{Simplicity}: $\bd{f}(\bar{\bd{x}}_j)$ should be computationally cheap to evaluate. It is especially important for PDE systems, where $\bd{f}$ may involve high-order derivatives.
    \item \textbf{Representativeness}: $\bar{\bd{x}}_j$ should cover the space of states where one aims to improve the model's performance or stability. Collocation points that a model might encounter and that are not represented by data snapshots are the best candidates.
    \item \textbf{Feasibility}: $\bar{\bd{x}}_j \in \XX$. In other words, $x_j$ should be an attainable state of the system. Collocation points outside of $\XX$ may downgrade the performance of the autoencoder by forcing it to be an invertible function on a domain outside of $\XX$.
\end{enumerate}
Thus, an optimal sampling procedure for collocation points $\bd{\bar{x}}_j$ is domain-specific and should be designed given a particular system $\bd{f}$ and available data $\bd{x}_i$. We show examples of how these conditions can be implemented for real systems in the following sections.

The above definition of collocation points is not to be confused with a classic notion of collocation points for finding numerical solutions for differential equations~\cite{fornberg1998practical, trefethen2022numerical}. The classic notion refers to a set of points in time $[t_0, t_0 + c_1h, t_0 + c_2h, \dots, t_0 + h]$, $0 < c1 < c2 < \dots < 1$ which are chosen to obtain an optimal local interpolant of a solution of a differential equation for a time-period between $t_0$ and $t_0 + h$. For example, $s$ collocation points for Runge-Kutta methods are defined to provide an optimal Gauss-Legendre interpolant of order $s$; the coefficients $c_1, \dots, c_s$ come from a respective Butcher table. In contrast, we define collocation points as pairs $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$ which are  examples of mapping $x \to f(x)$. Our definition is built around solving an \textit{inverse} problem of approximating $\dot{x} = f(x)$ with $f_\theta(x)$ and follows a recent work~\cite{liu2022physics} which develops upon a definition from \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.

\paragraph{Combined Loss Function} We train the model by optimizing a sum of the physics-informed loss~(\ref{eq:loss_physics_informed}) and the data-driven loss~(\ref{eq:loss_data_driven}):
\begin{equation}
    \label{eq:loss_combined}
    \min_\theta \left[\LL^{physics}_\theta + \LL^{data}_\theta\right]  
\end{equation}
When $\omega_1 = \omega_2 = 0$ we have $\LL^{data}_\theta = 0$, so we say that the model is (purely) \textit{\textbf{Physics-Informed}}. Similarly, when $\omega_3 = \omega_4 = 0$ we have $\LL^{physics}_\theta = 0$ and we say that the model is (purely) \textit{\textbf{Data-Driven}}. When $\omega_i \neq 0, \, \forall i$, we say that the model is \textit{\textbf{Hybrid}}.
